---
title: "Machine Learning in R: Workshop Series"
subtitle: "Introduction to the Tidyverse"
author: "Simon Schölzel"
institute: "*Research Team Berens*"
date: "2020-11-09 (updated: `r Sys.Date()`)"
output: html_notebook
---

This notebook complements the "**Introduction to the Tidyverse**" workshop which is part of the "Machine Learning in `R`" graduate course held at University of Münster, School of Business and Economics (winter term 2020/21).

For the purpose of reproducibility, it contains all examples and use cases discussed during the workshop sessions.


## Package Management

```{r}
#check if pacman is installed (install if evaluates to FALSE)
if (!require(pacman) == T) install.packages("pacman")
#load (or install if pacman cannot find an existing installation) the relevant packages
pacman::p_load(tidyverse, plotly, patchwork, ggforce)
pacman::p_load_gh("allisonhorst/palmerpenguins")
```


## `palmerpenguins` Data Set

```{r}
penguins
```

## `magrittr`: A Forward-Pipe Operator for R

```{r}
mean(subset(penguins, year == 2007)$body_mass_g, na.rm = T)

#alternatively:
peng_bm_2007 <- subset(penguins, year == 2007)$body_mass_g
mean(peng_bm_2007, na.rm = T)
```
```{r}
penguins %>% 
  subset(year == 2007) %>% 
  .$body_mass_g %>% 
  mean(na.rm = T)
```


## `tibble`: Simple Data Frames

`tibble()`:
```{r, results=F}
tibble::tibble(
  x = c("a", "b"),
  y = c(1, 2),
  z = c(T, F)
)
```
`tribble()`:
```{r, results=F}
tibble::tribble(
  ~x, ~y,  ~z,
  "a", 1,  T,
  "b", 2,  F
)
```
`as_tibble()`:
```{r}
df <- data.frame(
  x = c("a", "b"),
  y = c(1, 2),
  z = c(T, F)
)

tibble::as_tibble(df)
```
`enframe()`:
```{r}
c(x = "a", y = "b", z = 1) %>%
  tibble::enframe(name = "x", value = "y")
```


## `readr`: Read Rectangular Text Data

`write_csv()`:
```{r}
penguins %>% 
  readr::write_csv(file = "./penguins.csv")
```
`read_csv()`:
```{r}
penguins <- readr::read_csv("./penguins.csv")
```
`read_csv()` with explicit column specifications:
```{r}
readr::read_csv(
  "./penguins.csv",
    col_types = cols(
      species = col_character(),
      year = col_datetime(format = "%Y"),
      island = col_skip()
    )
  )
```
`read_csv()` with changing the default for `guess_max`:
```{r}
readr::read_csv(file = "./penguins.csv", guess_max = 1001)
```


##`tidyr`: Tidy Messy Data

`pivot_longer()`:
```{r}
penguins_long <- penguins %>% 
  #create id column here to assign each observation a unique key
  dplyr::mutate(id = dplyr::row_number(), .before = species) %>% 
  tidyr::pivot_longer(
    cols = contains("_mm"),
    names_to = "meas_type", values_to = "measurement"
  )

penguins_long
```
`pivot_wider()`:
```{r}
penguins_long %>% 
  tidyr::pivot_wider(
    names_from = "meas_type",
    values_from = "measurement"
  )
```
`nest()`:
```{r}
nested_penguins <- penguins %>% 
  tidyr::nest(
    nested_data = c(island, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g, sex)
  )

nested_penguins
```
`unnest()`:
```{r}
nested_penguins %>% 
  tidyr::unnest(col = nested_data)
```
`unnest_wider()` to unpack columns: 
```{r}
nested_penguins %>% 
  tidyr::unnest_wider(col = nested_data)
```
`unnest_longer()` to unpack rows (here `island`): 
```{r}
nested_penguins %>% 
  tidyr::unnest_wider(col = nested_data) %>% 
  tidyr::unnest_longer(col = c(island))
```
`unite()`:
```{r}
united_penguins <- penguins %>% 
  tidyr::unite(col = "spec_gender", c(species, sex), sep = "_", remove = T)

united_penguins
```
`separate()`:
```{r}
united_penguins %>% 
  tidyr::separate(col = spec_gender, into = c("species", "sex"), sep = "_", remove = T)
```
`complete()` to make implicit `NA` explicit:
```{r}
incompl_penguins <- tibble::tibble(
  species = c(rep("Adelie", 2), rep("Gentoo", 1), rep("Chinstrap", 1)),
  year = c(2007, 2008, 2008, 2007),
  value = c(rnorm(3, mean = 50, sd = 15), NA)
)

incompl_penguins
```
```{r}
incompl_penguins %>% 
  tidyr::complete(
    species, year, fill = list(value = NA)
)
```
`drop_na()` to make explicit `NA` implicit:
```{r}
incompl_penguins %>% 
  tidyr::drop_na(value)
```
`fill()` to replace explicit `NA` with previous value:
```{r}
incompl_penguins %>% 
  tidyr::fill(value, .direction = "down")
```
`replace_na()` to replace explicit `NA` with column mean:
```{r}
incompl_penguins %>%
  tidyr::replace_na(replace = list(value = mean(.$value, na.rm = T)))
```


## `dplyr`: A Grammar of Data Manipulation

`filter()` to filter for rows that fulfill condition:
```{r}
penguins %>% 
  dplyr::filter(species == "Adelie")
```
```{r}
penguins %>% 
  dplyr::filter(is.na(bill_length_mm) == T)
```
```{r}
penguins %>% 
  dplyr::filter(between(body_mass_g, 3800, 4000) & (year < 2008 | year > 2008))
```
`slice()` to pick rows based on location:
```{r}
penguins %>% 
  dplyr::slice(23:26)
```
```{r}
penguins %>% 
  dplyr::slice_head(n = 5)
```
```{r}
penguins %>% 
  dplyr::slice_sample(prop = 0.02)
```
```{r}
penguins %>% 
  dplyr::slice_min(flipper_length_mm, n = 5)
```
`arrange()` to change the order of rows:
```{r}
penguins %>% 
  dplyr::arrange(body_mass_g) %>% 
  dplyr::slice_head(n = 3)
```
```{r}
penguins %>% 
  dplyr::arrange(desc(body_mass_g)) %>% 
  dplyr::slice_head(n = 3)
```

--- End of Pt. 1 ---
