---
title: "Machine Learning in R: Workshop Series"
subtitle: "Introduction to the Tidyverse"
author: "Simon Schölzel"
institute: "*Research Team Berens*"
date: "2020-08-20 (updated: `r Sys.Date()`)"

output: html_notebook
---

This notebook complements the "**Introduction to the Tidyverse**" workshop which is part of the Machine Learning in `R` course (winter term 2020/21). For the purpose of reproducibility, it contains all examples and use cases discussed in the workshop.


## Package Management

```{r}
#check if pacman is installed (install if evaluates to FALSE)
if (!require(pacman) == T) install.packages("pacman")
#load (or install if pacman cannot find an existing installation) the relevant packages
pacman::p_load(tidyverse, plotly, patchwork, ggforce)
pacman::p_load_gh("allisonhorst/palmerpenguins")
```


## `palmerpenguins` Data Set

```{r}
penguins
```

## `magrittr`: A Forward-Pipe Operator for R

```{r}
mean(subset(penguins, year == 2007)$body_mass_g, na.rm = T)

#alternatively:
peng_bm_2007 <- subset(penguins, year == 2007)$body_mass_g
mean(peng_bm_2007, na.rm = T)
```
```{r}
penguins %>% 
  subset(year == 2007) %>% 
  .$body_mass_g %>% 
  mean(na.rm = T)
```


## `tibble`: Simple Data Frames

`tibble()`:
```{r, results=F}
tibble::tibble(
  x = c("a", "b"),
  y = c(1, 2),
  z = c(T, F)
)
```
`tribble()`:
```{r, results=F}
tibble::tribble(
  ~x, ~y,  ~z,
  "a", 1,  T,
  "b", 2,  F
)
```
`as_tibble()`:
```{r}
df <- data.frame(
  x = c("a", "b"), y = c(1, 2), z = c(T, F)
)

tibble::as_tibble(df)
```
`enframe()`:
```{r}
c(x = "a", y = "b") %>%
  tibble::enframe(name = "x", value = "y")
```


## `readr`: Read Rectangular Text Data

`write_csv()`:
```{r}
penguins %>% 
  write_csv(file = "./penguins.csv")
```
`read_csv()`:
```{r}
penguins <- readr::read_csv("./penguins.csv")
```
`read_csv()` with explicit column specifications:
```{r}
readr::read_csv(
  "./penguins.csv",
    col_types = cols(
      species = col_character(),
      year = col_datetime(format = "%Y"),
      island = col_skip()
    )
  )
```
`read_csv()` with changing the default for `guess_max`:
```{r}
readr::read_csv(file = "./penguins.csv", guess_max = 1001)
```


##`tidyr`: Tidy Messy Data

`pivot_longer()`:
```{r}
penguins_long <- penguins %>% 
  #create id column here to assign each observation a unique key
  mutate(id = dplyr::row_number(), .before = species) %>% 
  tidyr::pivot_longer(
    cols = contains("_mm"),
    names_to = "meas_type", values_to = "measurement"
  )

penguins_long
```
`pivot_wider()`:
```{r}
penguins_long %>% 
  tidyr::pivot_wider(
    names_from = "meas_type", values_from = "measurement"
  )
```
`nest()`:
```{r}
nested_penguins <- penguins %>% 
  tidyr::nest(
    nested_data = c(island, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g, sex)
  )

nested_penguins
```
`unnest()`:
```{r}
nested_penguins %>% 
  unnest(col = nested_data)
```
`unnest_wider()` to unpack columns: 
```{r}
nested_penguins %>% 
  unnest_wider(col = nested_data)
```
`unnest_longer()` to unpack rows (here `island`): 
```{r}
nested_penguins %>% 
  unnest_wider(col = nested_data) %>% 
  unnest_longer(col = c(island))
```
`unite()`:
```{r}
united_penguins <- penguins %>% 
  tidyr::unite(col = "spec_gender", c(species, sex), sep = "_", remove = T)

united_penguins
```
`separate()`:
```{r}
united_penguins %>% 
  tidyr::separate(col = spec_gender, into = c("species", "sex"), sep = "_", remove = T)
```
`complete()` to make implicit `NA` explicit:
```{r}
incompl_penguins <- tibble(
  species = c(rep("Adelie", 2), rep("Gentoo", 1), rep("Chinstrap", 1)),
  year = c(2007, 2008, 2008, 2007),
  value = c(rnorm(3, mean = 50, sd = 15), NA)
)

incompl_penguins
```
```{r}
incompl_penguins %>% 
  tidyr::complete(
    species, year, fill = list(value = NA)
)
```
`drop_na()` to make explicit `NA` implicit:
```{r}
incompl_penguins %>% 
  drop_na(value)
```
`fill()` to replace explicit `NA` with previous value:
```{r}
incompl_penguins %>% 
  tidyr::fill(value, .direction = "down")
```
`replace_na()` to replace explicit `NA` with column mean:
```{r}
incompl_penguins %>%
  tidyr::replace_na(replace = list(value = mean(.$value, na.rm = T)))
```


## `dplyr`: A Grammar of Data Manipulation

`filter()` to filter for rows that fulfill condition:
```{r}
penguins %>% 
  filter(species == "Adelie")
```
```{r}
penguins %>% 
  filter(is.na(bill_length_mm) == T)
```
```{r}
penguins %>% 
  filter(between(body_mass_g, 3800, 4000) & year < 2008)
```
`slice()` to pick rows based on location:
```{r}
penguins %>% 
  slice(23:26)
```
```{r}
penguins %>% 
  slice_head(n = 5)
```
```{r}
penguins %>% 
  slice_sample(prop = 0.02)
```
```{r}
penguins %>% 
  slice_min(flipper_length_mm, n = 5)
```
`arrange()` to change the order of rows:
```{r}
penguins %>% 
  arrange(body_mass_g) %>% 
  slice_head(n = 3)
```
```{r}
penguins %>% 
  arrange(desc(body_mass_g)) %>% 
  slice_head(n = 3)
```
`select()` to pick respectively drop certain columns:
```{r}
penguins %>% 
  select(1:3)
```
```{r}
penguins %>% 
  select(species, island, bill_length_mm)
```
```{r}
penguins %>% 
  select(starts_with("s"))
```
```{r}
penguins %>% 
  select(ends_with("mm"))
```
```{r}
penguins %>% 
  select(contains("mm"))
```
```{r}
penguins %>% 
  select(-contains("mm"))
```
```{r}
penguins %>% 
  select(where(is.numeric)) %>%   #equivalent to select(where(~is.numeric(.)))
  select(where(~mean(., na.rm=T) > 1000))
```
`rename()` to change column names:
```{r}
penguins %>% 
  rename(bm = body_mass_g, gender = sex) %>% 
  colnames()
```
```{r}
penguins %>% 
  rename_with(.fn = toupper, .cols = contains("mm")) %>% 
  colnames()
```
`relocate()` to change the order of columns:
```{r}
penguins %>% 
  relocate(species, .after = body_mass_g) %>%
  relocate(sex, .before = species) %>%
  relocate(island, .after = last_col())
```
`mutate()` to change the values of columns and/or create new columns:
```{r}
penguins %>% 
  mutate(
    bm_kg = body_mass_g / 1000,
    .keep = "all",
    .after = body_mass_g)
```
```{r}
penguins %>% 
  mutate(
    sex = case_when(
      sex == "male" ~ 1,
      sex == "female" ~ 0),
    .keep = "all")
```
```{r}
penguins %>% 
  mutate(
    across(contains("mm"), ~ . / 1000),
    .keep = "all")
```
`group_by()` to group rows based on a set of columns:
```{r}
penguins %>% 
  group_by(species)
```
`summarise()` to reduce a group into a single row:
```{r}
penguins %>% 
  group_by(species) %>% #univariate
  summarise(count = n(), .groups = "drop")
```
```{r}
penguins %>% 
  group_by(species, year) %>% #bivariate
  summarise(count = n(), .groups = "drop")
```
```{r}
penguins %>% 
  group_by(species) %>%
  summarise(
    across(contains("mm"), ~mean(., na.rm = T), .names = "{.col}_avg"),
    .groups = "drop")
```
```{r}
penguins %>% 
  group_by(species) %>% 
  group_by(year, .add = T) 
```
```{r}
penguins %>% 
  group_by(species) %>%
  summarise(
    across(
      contains("mm"),
      list(avg = ~mean(., na.rm = T), sd = ~sd(., na.rm = T)),
      .names = "{.col}_{.fn}"),
    .groups = "drop")
```
```{r}
penguins %>% 
  group_by(species) %>% 
  mutate(stand_bm = (body_mass_g - mean(body_mass_g, na.rm = TRUE)) / 
           sd(body_mass_g, na.rm = TRUE))
```
```{r}
bm_breaks <- mean(penguins$body_mass_g, na.rm = T) -
  (-3:3) * sd(penguins$body_mass_g, na.rm = T)

penguins %>% 
  group_by(species, bm_cat = cut(body_mass_g, breaks = bm_breaks)) %>% 
  summarise(count = n(), .groups = "drop")
```
```{r}
penguins %>% 
  group_by(species, island) %>% 
  filter(flipper_length_mm == max(flipper_length_mm, na.rm = T))
```
```{r}
penguins %>% 
  group_by(species, year) %>% 
  nest
```
`distinct()` to select only unique rows:
```{r}
penguins %>% 
  distinct(species, island)
```
`pull()` to extract single columns as vectors:
```{r}
penguins %>% 
  pull(year) #equivalent to penguins$year
```
`if_else()` to apply a vectorized if-else-statement
```{r}
penguins %>% 
  select(species, island, body_mass_g) %>% 
  mutate(penguin_size = if_else(body_mass_g < 3500, "tiny penguin", "big penguin"))
```
`lag()` and `lead()` to shift column values by an offset `n`:
```{r}
penguins %>% 
  select(species, body_mass_g) %>% 
  mutate(
    lagged_bm = lag(body_mass_g, n = 1),
    lead_bm = lead(body_mass_g, n = 2))
```


## `purrr`: Functional Programming Tools

`map()` to apply a function to each element of a vector:
```{r}
z_transform <- function(.x) {
  mean <- mean(.x, na.rm = T)
  sd <- sd(.x, na.rm = T)
  return( (.x - mean) / sd )
}

penguins %>% 
  select(contains("mm")) %>% 
  map(.x = ., .f = ~z_transform(.x))
```
```{r}
penguins %>% 
  select(contains("mm")) %>% 
  map(.x = ., .f = function(.x) { (.x - mean(.x, na.rm = T)) / sd(.x, na.rm = T) })
```
```{r}
penguins %>% 
  select(contains("mm")) %>% 
  map(.x = ., .f = ~(.x - mean(.x, na.rm = T)) / sd(.x, na.rm = T))
```
```{r}
penguins %>%
  map_df(class)
```
```{r}
penguins %>%
  map_df(~sum(is.na(.)))
```
```{r}
penguins %>%
  map_df(n_distinct)
```
```{r}
penguins %>%
  drop_na %>% 
  group_by(sex) %>%
  group_map(~slice_max(., flipper_length_mm, n = 1), .keep = T)
```
```{r}
species <- penguins %>% distinct(species, year) %>% pull(species) #.x argument for map()
years <- penguins %>% distinct(species, year) %>% pull(year)      #.y argument for map()

map2(
  .x = species,
  .y = years,
  .f = ~{
    penguins %>%
      drop_na %>% 
      filter(species == .x, year == .y) %>% 
      ggplot() +
        geom_point(aes(x = bill_length_mm, y = body_mass_g)) +
        labs(title = glue::glue("Scatter Plot Bill Length vs. Body Mass ({.x}, {.y})"))
    })
```
```{r}
penguins %>% 
  drop_na %>% 
  group_by(species, island) %>% 
  nest %>% 
  mutate(lin_reg = map(.x = data, .f = ~lm(body_mass_g ~ ., data = .x)))  %>% 
  mutate(coefs = map(lin_reg, ~summary(.x) %>% .$coefficients %>% as_tibble)) %>%
  select(-data, -lin_reg) %>% 
  unnest(coefs)
```
`possibly()` and `safely()` to capture error messages
```{r}
possibly_lm <- possibly(.f = lm, otherwise = "Error message")

penguins %>% 
  mutate(across(c(sex, island), as.factor)) %>% 
  group_by(species) %>% 
  group_map(.f = ~possibly_lm(flipper_length_mm ~ bill_length_mm + island, data = .)) 
  #keep(., . == "Error message") #discard(., . == "Error message")
```
```{r}
safely_lm <- safely(.f = lm, otherwise = NULL)

penguins %>% 
  mutate(across(c(sex, island), as.factor)) %>% 
  group_by(species) %>% 
  group_map(.f = ~safely_lm(flipper_length_mm ~ bill_length_mm + island, data = .)) 
  #map(., "result") #map(., "error")
```


## `ggplot2`: Create Elegant Data Visualisations Using the Grammar of Graphics

**Univariate example:**
1) Add `data`:
```{r}
penguins %>% 
  ggplot(data = .) #equivalent to ggplot()
```
2) Add `aes`thetics mapping:
```{r}
penguins %>% 
  ggplot(
    aes(x = flipper_length_mm))
```
3) Add `geom`:
```{r}
penguins %>% 
  ggplot(aes(x = flipper_length_mm)) +
    geom_histogram(na.rm = TRUE)
```
```{r}
penguins %>% 
  ggplot(aes(x = flipper_length_mm)) +
    geom_bar(na.rm = TRUE) 
```
4) Add `stat`:
```{r}
penguins %>% 
  ggplot(aes(x = flipper_length_mm)) +
    geom_bar(stat = "density", na.rm = TRUE) 
```
Use `geom_density()` instead of `geom_bar(stat = "density")`:
```{r}
penguins %>% 
  ggplot(aes(x = flipper_length_mm)) +
    geom_density(na.rm = TRUE)
```
**Bivariate example:**
1) Add `data`:
```{r}
penguins %>% 
  ggplot()
```
2) Add `aes`thetics mapping:
```{r p_step_b2, eval=F}
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) 
```
3) Add `geom`:
```{r}
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
    geom_point(na.rm = TRUE)
```
```{r}
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
    geom_point(aes(color = species), na.rm = TRUE) 
```
```{r}
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
    geom_point(aes(color = bill_depth_mm), na.rm = TRUE) 
```
```{r}
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
    geom_point(color = "red", na.rm = TRUE) 
```
```{r}
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
    geom_point(aes(shape = species), size = 4, na.rm = TRUE) 
```
5) Add `facets`:
```{r}
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
    geom_point(aes(shape = species), na.rm = TRUE) +
    facet_wrap(~year)
```
```{r}
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
    geom_point(aes(shape = species), na.rm = TRUE) +
    facet_wrap(~year + island)
```
6) `scale` aesthetics mapping:
```{r}
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
    geom_point(aes(color = species), size = 3, na.rm = TRUE) +
    scale_colour_brewer(palette = "Set3") 
```
```{r}
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
    geom_point(aes(color = species), na.rm = TRUE) +
    scale_y_log10()  
```
**Other examples:**
Boxplots for numeric variables
```{r}
penguins %>% 
  pivot_longer(cols = contains("mm"), names_to = "var", values_to = "val") %>% 
  drop_na  %>% 
  ggplot(aes(x = var, y = val)) +
    geom_boxplot(na.rm = TRUE) +
    geom_jitter(alpha = 0.22, width = 0.3) 
```
Ordered bar chart
```{r}
penguins %>%
  dplyr::count(species) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  ggplot() +
    geom_col(aes(x = prop, y = forcats::fct_reorder(species, prop))) +
    scale_x_continuous(labels = scales::label_percent(1.))
```
Adjacent bar chart
```{r}
penguins %>% 
  ggplot(aes(x = species)) +
    geom_bar(aes(fill = island), position = "dodge")
```
Stacked bar chart
```{r}
penguins %>% 
  ggplot(aes(x = species)) +
    geom_bar(aes(fill = island), position = "stack")
```
```{r p_step_o6, eval=F}
penguins %>% 
  ggplot(aes(x = forcats::fct_lump(species, n = 1))) +
    geom_bar(aes(fill = island), position = "stack")
```
High-quality density plot
```{r}
p <- penguins %>% 
  ggplot(aes(x = body_mass_g)) +
    geom_density(aes(fill = species), na.rm = T, alpha = 0.4) +
    scale_x_continuous(breaks = seq(from = 3000, to = 6000, by = 500), limits = c(2000, 7000)) +
    scale_y_continuous(labels = scales::label_comma(accuracy = 0.0001)) +
    labs(
      title = "Density Function for Three Penguin Species of Palmer Penguins",
      subtitle = "Palmer Archipelago (2007-2009)",
      caption = "Data: https://github.com/allisonhorst/palmerpenguins",
      x = "Body mass [grams]",
      y = "Statistical density"
    ) +
    theme_classic() + #also: theme_minimal()
    theme(
      legend.position = "top",
      plot.title = element_text(size = 14, face = "bold"),
      plot.subtitle = element_text(size = 12),
      plot.caption = element_text(size = 10, face = "italic"),
      axis.text.x = element_text(size = 10),
      axis.text.y = element_blank(),
      axis.title = element_text(size = 10),
    )

p
```
Violin Plot
```{r}
penguins %>% 
  ggplot(aes(x = species, y = body_mass_g)) +
    geom_violin(aes(fill = species), na.rm = T) +
    theme_classic()
```
Lines of Best Fit
```{r}
penguins %>% 
  drop_na %>% 
  ggplot(aes(x = flipper_length_mm,
             y = body_mass_g)) +
    geom_point(aes(color = species)) +
    geom_smooth(method = "lm", se = T)
```


##`plotly`: Interactive Web Graphics
```{r, out.height='70%', out.width='100%'}
plotly::ggplotly(p)
```


##`patchwork`: The Composer of Plots

```{r, out.width='75%', out.height='75%', fig.retina=3, fig.align='center'}
library(patchwork)
p + p + p
```
```{r, out.width='75%', out.height='75%', fig.retina=3, fig.align='center'}
library(patchwork)
p + (p / p)
```


## `ggforce`: Accelerating `ggplot2`

```{r ggforce, eval=F}
penguins %>% 
  drop_na %>% 
  ggplot(aes(x = .panel_x, y = .panel_y, col = sex, fill = sex)) +
    geom_autopoint(alpha = 0.5) +
    geom_autohistogram(alpha = 0.5) +
    facet_matrix(rows = vars(species, island, body_mass_g, flipper_length_mm), switch = "both", layer.diag = 2) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90))
```
